Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-13 background/background.tex
Fix
New figure

Compared to mammals, zebrafish can regenerate spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches due to this ability. Experiments found that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al. \cite{ref:Tsarouchas} studied the the complex interactions happen at the lesion site. The inflammation affects the regeneration, as if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, are related the inflammation, and they are dynamically controlled by immunes cells, i.e. cytokines works as media to control the axonal regeneration. Cytokines are small proteins, work as messenger to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map can be drawn, according to the evidences found by Tsarouchas et al. \cite{ref:Tsarouchas}. After injury, a massive immune response was observed, with the increasing numbers of several kinds of cells: macrophage, neutrophil. This dynamical system consists of complex interactions between cells and cytokines, For example, il-1$\beta$ promotes macrophages' increase, and as a major source, more macrophages leads to more expression of tnf-$\alpha$ mRNA. Consequently as media, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration, where generally il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect; tnf-$\alpha$ promotes the regeneration all the time. The presence of the cytokines also affects the number of immune cells, while as sources, more cells lead to more expression of cytokines. Interactions between different kinds of cells are also considered in the system.

This dynamic system is suitable for mathematical modelling, where a set of equations can be written to represent the dynamics of different objects e.g. cells and cytokines. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions iniside the system, based on the proposed hypothesis \cite{ref:Tsarouchas}, and furthermore based our own hypothesis which may include different interactions.

The biological background of works in this dissertation is mostly related the the experimental evidences, quantitative data and proposed hypothesis in \cite{ref:Tsarouchas}. The published data on the quantitative measurement of cells and cytokines make the modelling and further analysis possible. Some other interactions proposed in our modelling stages are enlightened from Anderson et al. \cite{Anderson}.

>>>>

Compared to mammals, zebrafish can regenerate spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches due to this ability. Experiments found that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al.\cite{ref:Tsarouchas} studied the the complex interactions happen at the lesion site. The inflammation affects the regeneration, as if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration\cite{ref:Tsarouchas}. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, are related the inflammation, and they are dynamically controlled by immunes cells, i.e. immune cells produce and control the cytokines-mediated inflammation \cite{ref:Tsarouchas}. Cytokines are small proteins, work as messenger to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map (Figure \ref{fig:map}) can be drawn according to the evidences found in experiments.\cite{ref:Tsarouchas}. After injury, a massive immune response was observed, with the increasing numbers of several kinds of cells. This dynamical system consists of complex interactions and dynamical feedback between cells and cytokines, For example, il-1$\beta$ promotes macrophages' increase, and as a major source, more macrophages leads to more expression of tnf-$\alpha$ mRNA.

Consequently, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration. il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect; tnf-$\alpha$ promotes the regeneration all the time. The presence of the cytokines also affects the number of immune cells, while as sources, more immune cells produce more cytokines. Interactions between different types of cells are also considered in the system (e.g., the number of neutrophil is affected by macrophages).

\begin{figure}[t]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/network.png}}
    \end{center}

    \caption[Working model of the influence of the innate immune system on axonal regrowth]%
    {Working model of the influence of the innate immune system on axonal regrowth, from Tsarouchas et al.\cite{ref:Tsarouchas}. Lines ended with arrow represent promoting effect, lines ended with T-connectors represent inhibition}
    \label{fig:map}

This dynamic system is suitable for mathematical modelling, where a set of equations can be written to represent the dynamics of different objects e.g. each types of cells and cytokines. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions inside the system, based on the proposed hypothesis \cite{ref:Tsarouchas}, and furthermore based our own hypothesis which may include different interactions.

The biological background of works in this dissertation is mostly related the the experimental evidences, quantitative data and proposed hypothesis in \cite{ref:Tsarouchas}. The published data on the quantitative measurement of cells and cytokines make the modelling and further analysis possible. Some other interactions proposed in our modelling stages are enlightened from Anderson et al. \cite{Anderson}, where intracellular cytokine signaling networks were established and computationally modelled.

### 2020-08-13 introduction/introduction.tex
New figure

While we are interested in the tissue regeneration among different species, mathematical models and techniques can help us to resolve complex interactions and effects of cells and chemicals in a biological system. The objective in this dissertation is an example of complex biology system found in tissue regeneration: zebrafish spinal cord regeneration. Zebrafish has the ability to regenerate its spinal cord after injury; at the lesion site various immune cells are presented and complex interactions are found while they are regulating inflammation and repair.

Our interest lies in the modeling of the regeneration process that happens around the lesion site, especially in the model parameter and comparison of different models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was firstly proposed but the addressed interactions remains to be precisely defined, and quantitatively examined in experiments. However, we would like to explore the the `best fit' of the model which comes with a set of parameter estimates, given the observed data. Also, we wish to compare different it with other alternative models and possibly find the best model to describe the observed data. In these steps, models are expected to retain mathematical and biological significance, where differential equations are suitable to describe this system.

Fo the parameter estimation tasks, Bayesian inference techniques can help in the problems of over-fitting and local optima \cite{ref:abcsysbio}, compared to optimisation-based methods. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression for the dynamic system can be hard.

>>>>

% While we are interested in the tissue regeneration among different species, mathematical models and techniques can help us to resolve complex interactions and effects of cells and chemicals in a biological system.

% The objective in this dissertation is an example of complex biology system found in tissue regeneration: zebrafish spinal cord regeneration.

Zebrafish has the ability to regenerate its spinal cord after injury; tissue regeneration observed at the injury site is a complex biological process with interacting parts and dynamical feedback between different types of cells and molecules. These cells and molecules are involved in the inflammation and repair in a dynamical way and a complex interacting network can be drawn according to it.

% at the lesion site various immune cells are presented and complex interactions are found while they are regulating inflammation and repair.

% [think about what the appropriate context is to start the introduction with. You could for example talk about regeneration of tissues being a complex biological process with many interacting parts and dynamical feedback between different cell types and signalling molecules. Mathematical models can help to resolve this complexity. An important problem is how to parameterise such models from available data.

% or you could talk about mathematical models being useful in biology in general, then transition to the example of tissue regeneration.]

Mathematical models and techniques can help us to resolve the complexity in this dynamic system, however, how to derive the parameter estimates of the models remains a challenging work. In this dissertation, our interest lies in the modeling of the regeneration process that happens around the lesion site, especially in the parameter estimates and comparison of possible models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was firstly proposed but the addressed interactions remains to be precisely defined, and quantitatively examined in experiments. We would like to find the best parameter sets of the model using the published data, and ranking models under certain metrics and possibly find the best-fit model. In these steps, models are expected to retain mathematical and biological significance regarding the complex interactions and differential equations are suitable for this.

% Also, we wish to compare different it with other alternative models and possibly find the best model to describe the observed data.

Fo the parameter estimation tasks, optimisation-based methods usually have trouble in determining global optima from local optimal, and noise in the observed data may lead to over-fitting. Compared to optimisation-based methods, Bayesian inference techniques can help to relief these two concerns\cite{ref:abcsysbio}. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression for the dynamic system can be hard.

### 2020-08-13 modelling/modelling.tex
New figure

\subsection{Model evaluation and comparison}

[general topics on evaluating a model]

[bayes factor for model selection]

@@

[hypothesis]

[available data]

[model misspecification]

>>>>

% \subsection{Model evaluation and comparison}

% [general topics on evaluating a model]

% There exists several metrics to evaluate the model based on the observed data. A widely used method is to simulate the data from the model, then calculate the discrepancy between simulated data and observed data. P-norm distance functions are usually used in this method

% [bayes factor for model selection]

@@

% [hypothesis]

% [available data]

% [model misspecification]

Some limitations were included in Section 3.1, where the observed data have a relative high standard deviation at certain measurement points; more measurement repeats or more measurement points could help to represent more accurate trajectories of the modelled variables and thus more possible models can be proposed. On the other hand, more accurate observed are alway beneficial to mathematical modelling, especially to complex dynamic systems.

As stated, our models are largely based on the experimental evidence from Tsarouchas et al.\cite{ref:Tsarouchas}. Some hypothesised interactions may differ from real case, and there may exist undiscovered mechanisms. These could lead to model misspecification, where the model could fit the data but with wrong interaction paths, or the model can hardly fit the data and recover some local features observed in the data.
